---
title: "INFSCI 2595 Fall 2020: Office hours"
subtitle: "Matrix math and debugging"
author: "Dr. Joseph P. Yurko"
date: "9/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, load_packages}
library(tidyverse)
```

Make a matrix to represent a design matrix.  

```{r, make_a_matrix}
set.seed(71231)
X <- cbind(rep(1, 7), rnorm(7))

colnames(X) <- c("(Intercept)", "x1")
```

```{r}
X
```

Make the coefficient vector.  

```{r}
beta_vec <- c(1.2, 2.5)

class(beta_vec)

length(beta_vec)

dim(beta_vec)
```

To convert the "regular" vector to a matrix data type we can use either `as.matrix()` or `matrix()`.  

```{r}
as.matrix(beta_vec)

class(as.matrix(beta_vec))
```

```{r}
matrix(beta_vec)

class(matrix(beta_vec))
```

We could force 1 row and 2 columns by:  

```{r}
matrix(beta_vec, nrow = 1)
```

We want to calculate the mean trend vector using matrix math:  

$$ 
\boldsymbol{\mu} = \mathbf{X} \boldsymbol{\beta}
$$

The design matrix has $N$ rows and $D+1$ columns. With a single input, $D=1$, the design matrix has 2 columns. To matrix multiply $\mathbf{X}$ and $\boldsymbol{\beta}$ the coefficient column vector must be 2 rows by 1 column.  

```{r}
X %*% matrix(beta_vec)
```

What happens if we use the $\boldsymbol{\beta}$ vector as a row vector?  

```{r}
X %*% matrix(beta_vec, nrow = 1)
```

What would happen if we tried $\boldsymbol{\beta} \mathbf{X}$?  

```{r}
matrix(beta_vec) %*% X
```

```{r}
matrix(beta_vec)
```

What happens if we did not convert from a regular vector a matrix and we used `*` instead of `%*%`?  

```{r}
beta_vec * X
```

What happens if we had the design matrix first?  

```{r}
X * beta_vec
```

So these are incorrect, because we should have 1 and only 1 mean value, $\mu_n$, per observation.  

```{r}
beta_vec * X[1, ]
```

```{r}
beta_vec * X[7, ]
```

Element wise multiplication takes each element of `beta_vec` and multiplies it by the corresponding column in `X` and repeats that action for every row in `X`.  

That is NOT what we want, because we want the INNER PRODUCT of each row of the design matrix with the coefficient vector.  

```{r}
as.vector(beta_vec[1] * X[1, 1] + beta_vec[2] * X[1, 2])
```

Check with the first row of the mean trend vector calculated by the matrix math.  

```{r}
X %*% matrix(beta_vec)
```

Check again with the second observation.  

```{r}
as.vector(beta_vec[1] * X[2, 1] + beta_vec[2] * X[2, 2])
```

This is different than, multiplying every element in the design matrix, $\mathbf{X}$, by the slope, $\beta_1$ (`beta_vec[1]`) and adding in the intercept, as shown below:  

```{r}
beta_vec[1] + beta_vec[2] * X
```

Let's see how we can debug our functions that use matrix multiplication.  

```{r}
my_function <- function(unknowns, my_info)
{
  Xmat <- my_info$the_matrix_x
  
  # enter debug mode with browser()
  # browser()
  
  # force a return as a "hack" to debugging
  # return( dim(unknowns) )
  
  # first attempt, is this correct?  does the fucntion crash?
  # mu <- unknowns[1] + unknowns[2] * Xmat
  
  # second attempt, is this correct? does the function crash?
  mu <- unknowns %*% Xmat
  
  # by default the very last line in a function is returned
  sum(dnorm(x = my_info$yobs,
            mean = mu,
            sd = my_info$sd_use))
}
```

Create the list of required information by first generating some random responses around the mean.  

```{r}
set.seed(61123)
y <- rnorm(n = nrow(X), mean = as.vector(X %*% matrix(beta_vec)), sd = 1)
```

```{r}
info_use <- list(
  the_matrix_x = X,
  yobs = y,
  sd_use = 1
)
```

Call our function, are we correct? Do we have an error?    

```{r}
my_function(beta_vec, info_use)
```

Use the debugger, via the `browser()` function. Uncomment out the line with `browser()` in the `my_function()` function and then rerun the code chunk to assign the modified function to `my_function()`. Run the code chunk above to launch debug mode! Within debug mode, RStudio will open a new Tab with a "glasses" icon to denote "inspection mode". You will be inside the `my_function()` workspace with access to the variables just as `my_function()` sees things. You can then test out operations in the console and use the Next/Run/Stop buttons to work proceed through the function. This is a useful way to test out errors inside functions.  

Alternatively, you can "hack" debugging by forcing returns. I like this approach when I just want to print out values to the screen as I build up the pieces of a function. Try commenting the line with `browser()` and uncomment the line with `return (dim(unknowns) )`, then rerun the code chunk to reassign `my_function()`. You will now have `my_function()` return the dimensions of the `unknowns` object, and then it exits without proceeding further.  